---
title: "dataone R package overview"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
<!-- output: pdf_document -->
vignette: >
  %\VignetteIndexEntry{dataone R Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## *dataone* R Package Overview

The *dataone* R package enables R scripts to search, download and upload science data and metadata
to the [DataONE Federation](https://www.dataone.org). This package calls DataONE 
web services that allow client programs to interact with [Member Nodes (MN)](https://www.dataone.org/current-member-nodes) 
and DataONE Coordinating Nodes (CN).

## Quick Start

See the full manual (`help dataone`) for documentation.

To search the DataONE Federation Member Node *Knowledge Network for Biocomplexity (KNB)* for a dataset:

```{r}
library(dataone)
cn <- CNode("PROD")
mn <- getMNode(cn, "urn:node:KNB")
mySearchTerms <- list(q="abstract:salmon+AND+keywords:acoustics+AND+keywords:\"Oncorhynchus nerka\"",
                      fl="id,title,dateUploaded,abstract,size",
                      fq="dateUploaded:[2013-01-01T00:00:00.000Z TO 2014-01-01T00:00:00.000Z]",
                      sort="dateUploaded+desc")
result <- query(mn, solrQuery=mySearchTerms, as="data.frame")
result[1,c("id", "title")]
pid <- result[1,'id']
```


The metadata file located in the above search can be downloaded with the commands:

```{r}
library(XML)
metadata <- rawToChar(getObject(mn, pid))
```

The metadata file that describes the located research can be viewed in an XML viewer or text editor, once
it is written to a disk file. This file details a data file (CSV) that can be obtained using the listed 
identifier, using the commands:

```{r}
dataRaw <- getObject(mn, "df35d.443.1")
dataChar <- rawToChar(dataRaw)
theData <- textConnection(dataChar)
df <- read.csv(theData, stringsAsFactors=FALSE)
df[1,]
```

Uploading a CSV file to a DataONE Member Node requires user authentication. DataONE user
authentication is described in the vignette `DataONE-Federation`.

Once the authentication steps have been followed, uploading is done with:
```{r}
library(datapack)
library(uuid)
d1c <- D1Client("STAGING", "urn:node:mnStageUCSB2")
id <- paste("urn:uuid:", UUIDgenerate(), sep="")
testdf <- data.frame(x=1:10,y=11:20)
csvfile <- paste(tempfile(), ".csv", sep="")
write.csv(testdf, csvfile, row.names=FALSE)
# Build a DataObject containing the csv, and upload it to the Member Node
d1Object <- new("DataObject", id, format="text/csv", filename=csvfile)
```
```{r, eval=FALSE}
uploadDataObject(d1c, d1Object, public=TRUE)
```

## Additional Resources

The `dataone` R package vignettes can be viewed using the R `vignette` comand,
for example `vignette("dataone-overview")`.

The `dataone` vignettes describe these topics:

- Searching for data in DataONE described in vignette *searching-dataone*. 

- Downloading data from DataONE described in vignette *download-data*

- Uploading data to DataONE is described in vignettes *upload-data* and *update-package*. 

- The DataONE Federation is described in the vignettes *DataONE-Federation*. 

## Acknowledgements
Work on this package was supported by:

- NSF-ABI grant #1262458 to C. Gries, M. Jones, and S. Collins.
- NSF-DATANET grants #0830944 and #1430508 to W. Michener, M. Jones, D. Vieglais, S. Allard and P. Cruse

Additional support was provided for working group collaboration by the National Center for Ecological Analysis and Synthesis, a Center funded by the University of California, Santa Barbara, and the State of California.
