---
title: "DataONE Federation"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DataONE Federation}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---


DataONE is a community driven project providing access to data across multiple member repositories, supporting enhanced search and discovery of Earth and environmental data.

DataONE comprises a distributed network of data centers, science networks or organizations. These organizations can expose their data within the DataONE network through the implementation of the DataONE Member Node service interface. In addition to scientific data, Member Nodes can provide computing resources, or services such as data replication, to the DataONE community.

## DataONE Service Interface

Access to DataONE data is provided by the DataONE service interface (aka REST API). DataONE maintains a set of 
user oriented tools called the [Investigator Toolkit](https://www.dataone.org/investigator-toolkit) (ITK) that 
use this service interface to search, download and 
upload data to DataONE Member Nodes. 

The `dataone` R package is one of the tools in the ITK.

If more detailed information about the DataONE Federation services is required than is provided by the *dataone* R package help, then in-depth documentation for the DataONE web services can be found at:

- Detailed overview [DataONE CN API](https://purl.dataone.org/architecture/apis/CN_APIs.html)
- Detailed overview [DataONE MN API](https://purl.dataone.org/architecture/apis/MN_APIs.html)

Please see the [DataONE Glossary](https://purl.dataone.org/architecture/glossary.html) for definitions of 
some terms used in this document that are used to describe DataONE services and architecture.

For additional overview information about the DataONE Federation, please https://www.dataone.org.

## New Features in DataONE Version 2.0
The following features are available via the `dataone` package, for member nodes that have implemented the DataONE version 2.0 API. 

### 1. Series Identifiers
Each data, metadata, and resource map object in DataONE has a unique identifier, refered to in DataONE documentation as a Persistent
Identifier (PID). A PID is associated with one object in DataONE and always refers to the same object, the same set of bytes that are 
stored on the DataONE network. 

A data or metadata object can be updated on a DataONE Member Node by using the R method dataone::updateObject(), so that 
a new version of the object becomes the active version that is discoverable in searches of DataONE. The older
version is still available if the PID is known, but this version will not show up in DataONE searches. In order
to download the new version, the new PID must be discovered and specified when the object is downloaded.

With *dataone* Version 2.0, an additional, optional identifier can be associated with an object, the Series Identifier (SID).
Using SIDs the most current version of an object can be obtained, without the need to determine the PID of the latest version.
For example, if the SID is specified when the object is downloaded, the most recent version of that object will be downloaded.

### 2. DataONE User Authentication Using Tokens
Uploading data to DataONE requires that a DataONE user identity be provided. In DataONE Version 1.x, the 
identity of a user was provided by an X.509 client certificate. DataONE Version 2.0
adds an additional method to provide identity information - an *authentication token*.
An authentication token is an encrypted text string that is provided by DataONE that
can be sent to DataONE services to prove a user's identity.

Authentication tokens can be used with Member Nodes that have been upgraded to the DataONE Version 2.0 Member Node API

The process of providing user identity information either via an X.509 certificate or via an authentication token is refered to as *authentication*.

Authentication tokens can be obtained from a user's DataONE account settings web page. To create an authentication token:

- Navigate to https://search.dataone.org
- Click *Sign in*, or *Sign up* if necessary
- Once signed in, move the cursor over the user name and select 'My profile' in the drop down menu.
- Click on the "Settings" tab.
- Click on "Authentication Token" in the menu below "Welcome"
- Click on the "Token for DataONE R" tab.
- Click "Renew authentication token" if the token you have been using has expired.
- Click on the "Copy" button below the text window to copy the authentication string to the paste buffer.
- Note the identity string and expiration date of the token.
- In the R console, paste the string which is similiar to this example:
```
  options(dataone_token = "eyJhbGciOiJSUzI1NiJ9...")
```
(For this example, the typically very long token value has been truncated for brevity, however the entire string
must be used for the token to be valid.)

- Note that this command can be entered in a user's ~/.Rprofile file so that this command is entered
when R is started.
- Remember that the console command must be re-entered with a new token value when the token expires

If the authentication token is defined as shown above, it will automatically be used when using methods
from the *dataone* R client.

<span style="color:red">
*The authentication token must be safegaurded like a password. Anyone with access to it can access content in DataONE as the user identity contained in the token. Care should be taken to not add this code to any published scripts or documents. This code will expire after a certain time period after which it will be necessary to obtain a new one.*
</span>

Note that the token shown above is for use in the DataONE production environment. If it is necessary to use authentication in 
a DataONE test environment, then the steps shown above should be followed by navigating to the following web page to generate
the token: https://search.test.dataone.org. If you generate a token from this web page, then the R option name
used should be "dataone_test_token", not "dataone_token", for example:

```
options(dataone_test_token = "eyJhbGciOiJSUzI1NiJ9...")
```
(For this example, the typically very long token value has been truncated for brevity, however the entire string
must be used for the token to be valid.)

For an explanation of DataONE environments, please see [DataONE environments](#environments).

Detailed, technical information about user identities and authentication in DataONE can be viewed at 

[DataONE Authentication](https://purl.dataone.org/architecture/design/Authentication.html)

You can check your token by entering the following R commands:
```{r, eval=F}
libary(dataone)
am <- AuthenticationManager()
getTokenInfo(am)
```

The most important column of the information returned by `getTokenInfo()` is the `expired` column. If a token
is expired, then it will not provide authentication and must be re-created from the user's profile page as
shown above.

### 3. Ability To Update System Metadata
Metadata is maintained by DataONE for each object that has been uploaded to it. This SystemMetadata for an object contains
information such as the access policy that determines the users that can read or update the data, the data's format type,
how many replicated copies of the data to create, etc.

Member Nodes that have been upgraded to DataONE Version 2.0 Member Node API now the have the ability to update the system metadata of a data object without having to update (replace) the data object itself. So for example, an object can be uploaded to DataONE without having 'public' read enabled (the data creator or
*rightsholder* and possibly a specified list of users could have access however). At a later date, the system metadata. could be updated to allow public read.

See ```help("updateSystemMetadata")``` for more information.

## <a name="environments"/> DataONE Environments

DataONE nodes are separated into several networks or *environments*. Each environment provides an isolated installations of the
DataONE services, such that nodes in one environment do not communicate with nodes in another. 

Currently DataONE uses the following environments: *production*, *staging*, *staging2*, *sandbox*, *sandbox2*, *dev* and *dev2*.
All environments except production are *test* environments that may have different version of software than the
released version of the DataONE infrastructure, and may experience service outages as required by the DataONE development team.

Most users will only need to use the production and staging2 environments. 

The production environment is the current working release of the DataONE infrastructure and is the environment that supports the operations necessary to fully implement the DataONE system. There is only one production environment.

The staging environment provides an installation of DataONE infrastructure that is a copy of the production environment. It 
is used to prepare for a new release of the infrastructure by testing the upgrade and software replacement procedures. 

The staging2 is a copy of the production environment and can be used by DataONE Member Nodes that are preparing 
to join the production environment. Staging2 usually has the released version of the DataONE infrastructure, and is more
stable that the other test environments.

The sandbox and sandbox2 environments offer more stable environments than the dev environments, and is intended to provide a more stable system where new features or alternative implementations may be evaluated within an environment that is close to a particular release of the DataONE infrastructure.

The dev and dev2 environments are intended for use by DataONE developers and are unstable, with software upgrades and 
service outages occurring as needed by the development staff.

The *dataone* R package uses the following abbreviations for the DataONE environments:

DataONE environment | *dataone* R package abbreviation | 
------------------- | -------------------------------- | 
production          | PROD                             | 
stage               | STAGING                          | 
stage 2             | STAGING2                         | 
sandbox             | SANDBOX                          | 
sandbox2            | SANDBOX2                         | 
dev                 | DEV                              | 
dev2                | DEV2                             | 

Each DataONE environment has a CN that maintains a registry of all MNs in that network. The CN can be queried for a list of 
all MNs in a network, for example, to see the MNs that are currently in the production environmemt:
```{r, eval=F}
cn <- CNode("PROD")
unlist(lapply(listNodes(cn), function(x) x@name))
```

To see the URL service endpoint for a CN, you can view the `endpoint` slot:

```{r,eval=F}
cn@endpoint
```

To see the member nodes that are currenly in the STAGE2 environment:
```{r, eval=F}
cn <- CNode("STAGING2")
unlist(lapply(listNodes(cn), function(x) x@name))
```
