---
title: "Known Issues"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Known Issues}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Known Issues With dataone R Package Version 2.0
### Error Using X.509 Certificates 

Using an X.509 certificate for DataONE authentication on certain versions of Mac OS X can cause the following error:

```
Error in curl::curl_fetch_memory(url, handle = handle) : 
Problem with the local SSL certificate 
```  

(Note that it is recommended that you use DataONE authentication tokens instead of an X.509 certificate if possible.)

Changes in the Mac OS X system libraries in OS X Mavericks have taken away support of these X.509 certificates.
A workaround to make these certificates usable on Mac OS X with R is to install a version of the *curl* R package
that supports these certificates. 

On Mac OS X, the libcurl library can be installed with either [Mac Ports](https://www.macports.org) package manager
or the [HomeBrew](http://brew.sh) package manager. The HomeBrew package manager can be significantly faster to install
but either one will work provided the directions shown below are followed.

You can check if you have MacPorts installed by entering the following command in a terminal window:

```
port version
```

#### Create new *curl* package using MacPorts
If MacPorts is being used on your system, the following commands can be entered to install a *curl* package
that can read the certificate and allow them to be used by the *dataone* package for authentication
to a DataONE node. In a terminal window enter the commands:

```
sudo port install curl
```

```{r, eval=F}
Sys.setenv(LIB_DIR="/opt/local/lib")
Sys.setenv(INCLUDE_DIR="/opt/local/include")
install.packages("curl", type="source")
library(curl)
library(dataone)

# Remove the environment variables as they are no longer needed.
Sys.setenv(LIB_DIR="")
Sys.setenv(INCLUDE_DIR="")
```

At this point you should be able to use X.509 Certificates.

#### Create new *curl* package using HomeBrew
The HomeBrew software can be installed with the following command entered at a terminal window:

```
ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)"
```

Once HomeBrew has been installed, you can get the required curl libraries by entering the command:

```
brew install curl --with-openssl
brew link curl --force
```

In the R console enter the commands:

```{r, eval=F}
Sys.setenv(LIB_DIR="/usr/local/opt/curl/lib")
Sys.setenv(INCLUDE_DIR="/usr/local/opt/curl/include")
install.packages("curl", type="source")
library(curl)
library(dataone)

# Remove the environment variables as they are no longer needed.
Sys.setenv(LIB_DIR="")
Sys.setenv(INCLUDE_DIR="")
```
At this point you should be able to use X.509 Certificates.
