---
title: "Updating Packages on DataONE"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uploading Data to DataONE}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Updating A DataONE Package

After a package has been uploaded to a DataONE Member Node, it may be determined by the package submitter or other interested
parties that the package needs to be updated, 
for example to add a missing file, replace one file with another, or remove a package member from the package.

These types of modifications can be accomplished by downloading the package from DataONE using the 
`getDataPackage` method to create a local copy of the package in R, then modifying the package contents and 
uploading the package back to the DataONE. 

### 1. Create A Package And Upload To DataONE

To demonstrate a package update, a simple example package will be used from files provided with the `dataone`
R package. This example package must first be created and then uploading to DataONE. Once this has been accomplished, the steps
necessary to update it will be shown.

First, a DataPackage is created and a sample EML metadata file is added.
This EML file contains descriptions of all the DataObjects that will be added to and uploaded with the example package:

```{r}
library(dataone)
library(datapack)
library(uuid)

d1c <- D1Client("STAGING", "urn:node:mnStageUCSB2")
dp <- new("DataPackage")

emlFile <- system.file("extdata/strix-pacific-northwest.xml", package="dataone")
metadataObj <- new("DataObject", format="eml://ecoinformatics.org/eml-2.1.1", filename=emlFile,
                   suggestedFilename=basename(emlFile))
dp <- addMember(dp, metadataObj)
```

Next, an input data file will be added to the package:

```{r}
sourceData <- system.file("extdata/sample.csv", package="dataone")
sourceObj <- new("DataObject", format="text/csv", filename=sourceData, suggestedFilename=basename(sourceData))
dp <- addMember(dp, sourceObj, metadataObj)
```

The next step shows how to update the DataObject that contains EML, directly within the DataPackage, rather than having
to extract the DataObject from the DataPackage, update it, then re-insert it.

The section of the metadata that describes the dataset `sample.csv` will be updated. This is an EML 
`<otherEntity>` element,  which contains the `<url>` element:
```
<physical>
    <distribution>
        <online>
            <url>some url value here</url>
        </online>
    </distribution>
</physical>
```
This element contains the link that can be used to download the file from DataONE, 
which includes the DataONE identifier for the object. Because the identifier value is set when the DataObject
is created, the EML metadata is out of date and needs to be updated with the new identifier value.

In the example below, the metadata is updated to have the distribution URL similiar to:
`https://mn-stage-ucsb-2.test.dataone.org/metacat/d1/mn/v2/object/urn:uuid:563c265b-3743-4e03-bcee-8ec09697a1a3`,
using the `updateMetadata` method:

```{r}
xpathToURL <- "//otherEntity/physical/distribution[../objectName/text()=\"sample.csv\"]/online/url"
newURL <- sprintf("%s/%s/object/%s", d1c@mn@baseURL, d1c@mn@APIversion, getIdentifier(sourceObj))
dp <- updateMetadata(dp, do=metadataObj, xpath=xpathToURL, replacement=newURL)
```

The `updateMetadata` method updates the DataObject specified with the `do` argument. This argument can alternatively
be specified as the identifier value of the DataObject. 

The `xpath` argument specifies a location in the EML document as an XML Xpath expression. The `replacement`
argument specifies the character string that will be substituted for the previous value. An updated DataPackage
containing the updated metadata DataObject is returned from the call.

Also, `updateMedata` will update the identifier associated with the DataObject if necessary.
and retain the original identifier. The original identifer and a new identifier 
are required by DataONE when the modified object is uploaded to DataONE. 

If the `newId` parameter  is not included in the call to `updateMetadata`, then the new identifier will be generated automatically. Alternatively, the `newId` parameter can be used to specify a desired value for the new identifier

The identifier for the DataObject can be retrieved from a DataPackage using the `selectMember` method:
```{r}
metadataId <- selectMember(dp, name="sysmeta@formatId", value="eml://ecoinformatics.org/eml-2.1.1")
```

The `selectMember` method checks the `name` slot, in this case `sysmeta@formatId`, of each DataObject in the 
DataPackage `dp` and returns the identifier of any member that matches the specified 
value (`eml://ecoinformatics.org/eml-2.1.1` in this case). Only the DataObject that contains the EML metadata i
matches for the example DataPackage, so the identifier for just that member is returned.

Next, an R script and data file will be added to the DataPackage in the same manner as the previous file.
Note that the identifier for the metadata DataObject can be used instead of the DataObject itself for calls 
to `addMember` (argument `mo`) and `updateMetadata` (argument `do`):
```{r}

progFile <- system.file("extdata/filterSpecies.R", package="dataone")
progObj <- new("DataObject", format="application/R", filename=progFile, mediaType="text/x-rsrc", suggestedFilename=basename(progFile))
dp <- addMember(dp, progObj, mo=metadataId)

xpathToURL <- "//otherEntity/physical/distribution[../objectName/text()=\"filterSpecies.R\"]/online/url"
newURL <- sprintf("%s/%s/object/%s", d1c@mn@baseURL, d1c@mn@APIversion, getIdentifier(progObj))
dp <- updateMetadata(dp, metadataId, xpath=xpathToURL, replacement=newURL)
metadataId <- selectMember(dp, name="sysmeta@formatId", value="eml://ecoinformatics.org/eml-2.1.1")

outputData <- system.file("extdata/filteredSpecies.csv", package="dataone")
outputObj <- new("DataObject", format="text/csv", filename=outputData, suggestedFilename=basename(outputData))
dp <- addMember(dp, outputObj, mo=metadataId)

xpathToURL <- "//otherEntity/physical/distribution[../objectName/text()=\"filteredSpecies.csv\"]/online/url"
newURL <- sprintf("%s/%s/object/%s", d1c@mn@baseURL, d1c@mn@APIversion, getIdentifier(outputObj))
dp <- updateMetadata(dp, do=metadataId, xpath=xpathToURL, replacement=newURL)
```
Now upload the package to DataONE:
```{r, eval=FALSE}
packageId <- uploadDataPackage(d1c, dp, public=TRUE, quiet=FALSE)

```
Note that uploading a package to DataONE requires valid DataONE user authentication.

### 2. Download The Package From DataONE

The first step in updating a package is to download the package from DataONE so that it can 
be modified using methods in the `dataone` package. Modifications can be made to the package such as adding 
or removing members from the package, or changing the contents of a package member, as would be the case 
if the wrong file was initially uploaded.

Because calling `uploadDataPackage` to upload a package to DataONE requires valid DataONE user authentication,
this demonstration will use a locally saved DataPackage that was previously created using a valid DataONE account 
using the `uploadDataPackage` method shown in the previous section, then using the following statements:

```{r, eval=FALSE}
d1c <- D1Client("STAGING", "urn:node:mnStageUCSB2")
# The variable 'packageId' was obtained from the previous call to 'uploadDataPackage'
pkg <- getDataPackage(d1c, identifier=packageId, lazyLoad=TRUE, limit="0MB", quiet=FALSE)
saveRDS(pkg, file="./extdata/pkg.rds")
```
For this example, load the saved DataPackage object, which simulates the download that would occur
using `getDataPackage`.
```{r}
updatePkg <- readRDS(system.file("./extdata/pkg.rds", package="dataone"))
```

The `getDataPackage` method downloads all files belonging to the package specified by
the `identifier` parameter. 

Because packages might contain large files or a large number of files, it is
possible to `lazyLoad` the package. This means that the files are downloaded from the member node
only when needed, so initially the system metadata is downloaded but not the data itself. An upper
size limit can be specified when using `lazyLoad` by also specifying the `limit` parameter. Files that are larger than
the `limit` value are not downloaded, only system metadata. 

Note that metadata files, such as the EML in this example, are always downloaded regardless of the `lazyLoad` parameter value.

### 3. Review Package Contents.

The downloaded package can be viewed by typing the object name at the console, which invokes the
R `show` method for the object:
```{r, echo=FALSE}
saveWidth <- getOption("width")
options(width=100)
```
```{r}
updatePkg
```
```{r, echo=FALSE}
options(width=saveWidth)
```
Note that the `show` output for a DataPackage is condensed to fit the width of the current R console. If 
the output is condensed and more detail is required, set the R console with to a larger value (options=<width>),
or if using `Rstudio`, widen the console window.

The identifier for the metadata DataObject that describes the package members will be retrieved from the DataPackage:
```{r}
metadataId <- selectMember(updatePkg, name="sysmeta@formatId", value="eml://ecoinformatics.org/eml-2.1.1")
```

### 4. Modify DataObjects In The Package
The original uploaded package included the file `filterSpecies.csv`. This file can be substituted
for a different file, as would be necessary if it was determined that the zipped form of the
file should have been used.

First, determine which DataObject in the DataPackage `updatePkg` contains the errant file:
```{r}
objId <- selectMember(updatePkg, name="sysmeta@fileName", value='filteredSpecies.csv')
```

Next, update the DataObject in `updatePkg` to replace the file `filteredSpecies.csv` with `filteredSpecies.csv.zip` 
using the `replaceMember` method:
```{r}
zipfile <- system.file("extdata/filteredSpecies.csv.zip", package="dataone")
updatePkg <- replaceMember(updatePkg, objId, replacement=zipfile, formatId="application/octet-stream", 
                     suggestedFilename=basename(zipfile))
objId <- selectMember(updatePkg, name="sysmeta@fileName", value='filteredSpecies.csv.zip')
```

The `replaceMember` method replaces the data content of the DataObject, updates the relevant system metadata slots such as `size` and `checksum`. Similiar
to the `updateMetadata` method, if the `newId` parameter is not specified for `replaceMember`, then a new identifier
will be automatically generated if necessary.

Next, update the filename in the EML in the relevant section:
```{r}
nameXpath <- '//otherEntity/physical/objectName[text()="filteredSpecies.csv"]'
newName <- basename(zipfile)
updatePkg <- updateMetadata(updatePkg, metadataId, xpath=nameXpath, replacement=newName)
metadataId <- selectMember(updatePkg, name="sysmeta@formatId", value="eml://ecoinformatics.org/eml-2.1.1")

```
Since the identifier for the replace DataObject may have been updated, update the corresponding
distribution URL in the metadata:
```{r}
# Update the distribution URL with the new id from 'replaceMember'
urlXpath <- sprintf("//otherEntity/physical/distribution[../objectName/text()=\"%s\"]/online/url", basename(zipfile))
newURL <- sprintf("%s/%s/object/%s", d1c@mn@baseURL, d1c@mn@APIversion, objId)
updatePkg <- updateMetadata(updatePkg, metadataId, xpath=urlXpath, replacement=newURL)
metadataId <- selectMember(updatePkg, name="sysmeta@formatId", value="eml://ecoinformatics.org/eml-2.1.1")

```
In this example, the file `collectionMethods.csv` was mistakenly omitted from the original package upload,
so add it now:

```{r}
# Now add a new package member that was omitted from the original package
auxFile <- system.file("extdata/collectionMethods.csv", package="dataone")
auxObj <- new("DataObject", format="text/csv", filename=auxFile, suggestedFilename=basename(auxFile))
updatePkg <- addMember(updatePkg, auxObj, metadataId)

```
The metadata will be updated to have the correct download URL as shown for previous package members.

```{r}
urlXpath <- sprintf("//otherEntity/physical/distribution[../objectName/text()=\"%s\"]/online/url", auxObj@sysmeta@fileName)
newURL <- sprintf("%s/object/%s", d1c@mn@baseURL, getIdentifier(auxObj))
updatePkg <- updateMetadata(updatePkg, metadataId, xpath=urlXpath, replacement=newURL)
```

The package can be reviewed before updating:

```{r, echo=FALSE}
saveWidth <- getOption("width")
options(width=100)
```
```{r}
updatePkg
```
```{r, echo=FALSE}
options(width=saveWidth)
```

### 5. Upload The Modified DataPackage
Now upload the modified package to DataONE. Each DataObject in the DataPackage will be inspected by 
`uploadDataPackage` and DataObjects that have been modified will be updated and DataObjects that have been
added to the DataPackage will be uploaded. 
```{r, eval=FALSE}
newPackageId <- uploadDataPackage(d1c, updatePkg, public=TRUE, quiet=FALSE)
```
