---
title: "Uploading Data to DataONE"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uploading Data to DataONE}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Introduction
  
This document describes how to use the *dataone* R package to upload data to DataONE, and how to perform
maintenance operations on the data after it has been uploaded.

The *dataone* package provides methods to allow R scripts to interact with DataONE Coordinating Nodes (CN) and Member Nodes (MN). The *dataone* R package takes care of the details of calling the 
corresponding DataONE web service on a DataONE node. For example, the *dataone* `createObject` R method calls the DataONE web service [MNStorage.create()](https://purl.dataone.org/architecture/apis/MN_APIs.html#MNStorage.create) that uploads a dataset to a DataONE MN. 

Before uploading any data to a DataONE MN, it is necessary to obtain a DataONE user identity, and the means to provide that user 
identity when data is uploaded. The method that DataONE uses to achieve this is known as *user identity authentication*, and 
requires that an *authentication token*, which is a character string, is provided during upload. The process to obtain this
token is described in the *dataone-overview* vignette, in the section **New Authentication Mechanism**, which is viewable with the R command `vignette("dataone-overview")`. (Note: DataONE originally used X.509 certificates for authentication, which are still supported.)

## Uploading a data package using `uploadDataPackage`
Datasets and metadata can be uploaded individually or as a data package. Uploading a data package
will be described first and a workflow for preparing and uploading a data package using the `uploadDataPackage` method will be shown. 
A complete script that uses this workflow is shown here: 

```{r, eval=FALSE}
library(dataone)
library(datapack)
library(uuid)
dp <- new("DataPackage")
sampleData <- system.file("extdata/sample.csv", package="dataone")
# Create a unique identifier string for the data object in a standard format.
dataId  <- paste("urn:uuid:", UUIDgenerate(), sep="")
dataObj <- new("DataObject", id=dataId, format="text/csv", file=sampleData)
dataObj <- setPublicAccess(dataObj)
sampleEML <- system.file("extdata/sample-eml.xml", package="dataone")
# Create a unique id string for the data object in a standard format
# Alternatively DOI string could used using "generateIdentifier(mn, scheme="DOI")"
metadataId <- paste("urn:uuid:", UUIDgenerate(), sep="")
metadataObj <- new("DataObject", id=metadataId, format="eml://ecoinformatics.org/eml-2.1.1", file=sampleEML)
metadataObj <- setPublicAccess(metadataObj)
dp <- addData(dp, dataObj, metadataObj)
d1c <- D1Client("STAGING", "urn:node:mnStageUCSB2")
packageId <- uploadDataPackage(d1c, dp, replicate=TRUE, public=TRUE, numberReplicas=2)
```

The following sections describe each line of this script in detail.

### 1. Create a DataPackage object.
In order to use `uploadDataPackage`, it is necessary to prepare an R *DataPackage* object that will serve as a container for the set
of files  that will be included in the data package. The following commands load the required libraries and creates 
an empty *DataPackage* object that will be added to later:

```{r,eval=F}
library(dataone)
library(datapack)
library(uuid)
dp <- new("DataPackage")
```
When using the `uploadDataPackage` method, data structures that are required by DataONE are created, 
configured and uploaded automatically with the data package. These data structures include a 
[ResourceMap](https://purl.dataone.org/architecture/design/DataPackage.html#generating-resource-maps)
that describes the data package, and SystemMetadata objects that contain DataONE system information for each 
of the science datasets and associated science metadata. 

### 2. Prepare a metadata file that will describe the files in the data package
The next step is to prepare a metadata file that will describe the science datasets in the
data package. The most common metadata format used in the DataONE network is the
Ecological Metadata Langauage (EML). Other supported formats include FGDC and ISO 19113.
Additional information about EML is available at https://knb.ecoinformatics.org/#external//emlparser/docs/index.html.

Detailed directions regarding authoring metadata documents are outside the scope of this document.

### 3. Determine what access your data and metadata should have 
The levels of access available to objects in DataOne are "read", "write", and "changePermission".
The "read" permission allows a user the ability to view the content of a DataONE object.
The "write" permission allows a user the ability to change the content of an object via update services. 

Permissions are hierarchical, so write permission also includes read permission
The "changePermission" permission allows the ability to change the access policy for an object and includes 
both read and write permissions.

Each of these permissions can be granted to a single user, a group of users, or the special *public* user
which means all users.

Each object in DataONE can have one or more access rules that control the access of that object. The
complete set of access rules for an object is refered to as its access policy.

The next section shows how to apply the desired access rules to items that will be added to a data package
before upload.

### 4. Create a DataObject for each data file
A *DataObject* must be created for each metadata file and data file that will be included in the data package.
The *DataObject* maintains information about an object that will be needed by DataONE.

A *SystemMetadata* object will be created automatically and stored in each DataObject. The *SystemMetadata* object will
be used by DataONE to maintain low level information about the dataset, such as the access policy, the user identity
of the *rightsholder* (the user identity that can modify access the dataset), which Member 
Nodes it can be replicated to, etc. 

The example below creates a *DataObject* for a science dataset:

```{r,eval=F}
sampleData <- system.file("extdata/sample.csv", package="dataone")
dataId <- paste("urn:uuid:", UUIDgenerate(), sep="")
dataObj <- new("DataObject", id=dataId, format="text/csv", file=sampleData)
```

An optional *user* argument can be specified when creating a *DataObject*, which will be used to set the DataONE *submitter* and *rightsholder* of the dataset when it is uploaded. The rightsholder is granted all access priviledges 
to the object.

If *user* is not specified for a *DataObject*, then the submitter and rightsholder for an object will automatically be set, when the object is uploaded to DataONE, to the DataONE user that created the authentication token or X.509 certificate. 

Note that if the *id* argument is not specified, a unique identifier will automatically be created and assigned to the DataObject.

Access rules can be added to each *DataObject* after it has been created. Access rules can be added to grant permissions to a
single user. Access can also be granted to the *public* user, which means any and all users. For example, public read access
can be set using the `setPublicAccess` method:

```{r,eval=F}
dataObj <- setPublicAccess(dataObj)
```

Individual access rules to be added for a DataONE user identity can also be added to the access policy. 

Access rules are added to a *DataObject* using the `addAccessRule` method. The following access rule will grant
user 'Peter Smith' changePermission access to the dataset, which will take effect after it is uploaded and 
available on a DataONE MN:

```{r,eval=F}
accessRules <- data.frame(subject="CN=Peter Smith A10499,O=Google,C=US,DC=cilogon,DC=org", permission="changePermission") dataObj <- addAccessRule(dataObj, accessRules)
```
The value of the *subject* argument in the above example ("CN=Peter Smith A10499,O=Google,C=US,DC=cilogon,DC=org") is the string value of a typical DataONE user identity. DataONE
user identities and user authentication are described in section *A New Authentication Mechanism* 
in the vignette *dataone-overview* (to view this vignette, type this command in the R console: `vignette("dataone-overview")`)

### 5. Create a DataObject for the metadata file
When a *DataObject* is created, a unique identifier is generated if one is not specified. This automatically generated
identifier has the format "urn:uuid:<a unique string>", for example "urn:uuid:c3443142-6260-4ea5-aaa1-1114981e04ad".

The following command creates the *DataObject* for the science metadata, using an automatically generated identifier:

```{r,eval=F}
sampleEML <- system.file("extdata/sample-eml.xml", package="dataone")
metadataObj <- new("DataObject", format="eml://ecoinformatics.org/eml-2.1.1", file=sampleEML)
```

Alternatively, a Digital Object Identfier (DOI) may be assigned to the metadata *DataObject*, using the *generateIdentifier* method:

```{r, eval=FALSE}
cn <- CNode("STAGING")
mn <- getMNode(cn, "urn:node:mnStageUCSB2")
doi <- generateIdentifier(mn, "DOI")
metadataObj <- new("DataObject", id=doi, format="eml://ecoinformatics.org/eml-2.1.1", file=sampleEML)
```    

(Note that the example uses a DataONE test environment *STAGING*, and not the production environment. In 
order to create an valid DOI, a production DataONE Member Node that supports creating DOIs must be
used.)

The `generateIdentifier` method requests that the DataONE MN generate a properly formatted DOI. 

### 6. Add each DataObject to the DataPackage
The DataPackage object serves as a container for a set of data objects that will be uploaded to DataONE. The metadata 
*DataObject* and all science data *DataObjects* must be added to the DataPackage before calling `uploadDataPackage`.

Relationships between the objects in a DataPackage are stored in the ResourceMap which is stored in and maintained by the DataPackage.
One type of relationship that is stored is between the science metadata and the science datasets that are described by
it. In the [DataONE data package](https://purl.dataone.org/architecture/design/DataPackage.html) implementation, this
relationship is the [CITO](http://vocab.ox.ac.uk/cito) *documents* relationship that links the metadata object
to science objects.

This relationship between the science metadata and science data objects will be made automatically for each science 
data object as it is added to the DataPackage, if the metadata object is included when the science data object is
added.

Now add the metadata object to the DataPackage:

```{r, eval=F}
dp <- addData(dp, metadataObj)
```

Then specify the metadata object when each science data object is added, associating the metadata object with the
science object:

```{r, eval=F}
dp <- addData(dp, do = dataObj, mo = metadataObj)
```

If there were additional DataObjects to add to the package, they would be added to the DataPackage and 
associated with the metadata object as follows:

```
dp <- addData(dp, do = dataObj2, mo = metadataObj)
dp <- addData(dp, do = dataObj3, mo = metadataObj)
dp <- addData(dp, do = dataObj4, mo = metadataObj)
```
### 7. Upload the DataPackage

When all *DataObjects* have been added to the DataPackage, call the `uploadDataPackage` method to upload the 
entire DataPackage:

```{r, eval=FALSE}
d1c <- D1Client("STAGING", "urn:node:mnStageUCSB2")
packageId <- uploadDataPackage(d1c, dp, replicate=TRUE, numberReplicas=2)
message(sprintf("Uploaded data package with identifier: %s", packageId))
```

(Note that the example uses a DataONE test environment *STAGING*, and not the production environment.)

After *uploadDataPackage* has been called sucessfully, the data package can be viewed on the member node, searched for
using the DataONE search facility. Note that if objects in DataONE are not publicly readable, and the authenticated
user performing the search isn't granted access in an object's access policy, then the objects will not be 
viewable or discoverable via the search facility for that user.

## Uploading Individual Data And Metadata files

A single data or metadata file can be uploaded to a DataONE MN using the *createObject* method.
When uploading a single file using this method, additional information must be supplied to DataONE that
controls how DataONE interacts with the uploaded file. This additional information is stored
in DataONE as a *system metadata* object and contains information such as who can access or update
the file, how many copies of the file should be maintained, whether the file has been superseded by
another object, etc. The system metadata information that will be uploaded to DataONE is collected and stored in an R object type *datapack::SystemMetadata*, as shown below:

```{r,eval=F}
library(digest)
# Create a system metadata object for a data file. 
# Just for demonstration purposes, create a temporary data file.
testdf <- data.frame(x=1:20,y=11:30)
csvfile <- paste(tempfile(), ".csv", sep="")
write.csv(testdf, csvfile, row.names=FALSE)
format <- "text/csv"
size <- file.info(csvfile)$size
sha1 <- digest(csvfile, algo="sha1", serialize=FALSE, file=TRUE)
# Generate a unique identifier for the dataset
pid <- sprintf("urn:uuid:%s", UUIDgenerate())
sysmeta <- new("SystemMetadata", identifier=pid, formatId=format, size=size, checksum=sha1)
sysmeta <- addAccessRule(sysmeta, "public", "read")
```

Alternatively, the system metadata could have been created with a *seriesId*. The *seriesId* is explained in the
*dataone_overview* vignette. The following example shows the creation of a *SystemMetadata* object using the
optional *seriesId*:

```{r,eval=F}
# Create a system metadata object for a data file. 
# Just for demonstration purposes, create a temporary data file.
testdf <- data.frame(x=1:20,y=11:30)
csvfile <- paste(tempfile(), ".csv", sep="")
write.csv(testdf, csvfile, row.names=FALSE)
format <- "text/csv"
size <- file.info(csvfile)$size
sha1 <- digest(csvfile, algo="sha1", serialize=FALSE, file=TRUE)
# Generate a unique identifier for the dataset
pid <- sprintf("urn:uuid:%s", UUIDgenerate())
# The seriesId can be any unique character string.
seriesId <- sprintf("urn:uuid:%s", UUIDgenerate())
sysmeta <- new("SystemMetadata", identifier=pid, formatId=format, size=size, checksum=sha1,  seriesId=seriesId)
```
A unique identifier must be specified for each system metadata, whether or not a seriesId is used.

The dataset can now be uploaded to DataONE with the associated system metadata:
```{r,eval=F}
cn <- CNode("STAGING")
mn <- getMNode(cn, "urn:node:mnStageUCSB2")
response <- createObject(mn, pid, csvfile, sysmeta)
```

Note that for this example, the DataONE test environment *STAGING* is used, and not the production environment.

## Maintaining Uploaded Datasets
After data has been uploaded to DataONE, maintenance operations can be performed on these objects using the methods described in the following sections.

### Update the DataONE system metadata for an object (MNode: updateSystemMetadata)
The system metadata can be updated for an object in DataONE without updating the data bytes of the object itself. For
example, if an object was only readable by the data submitter, the access policy for an object can be updated to enable 
public access. System metadata is updated for an object using the *updateSystemMetadata* method:

The following exmaple first downloads the current system metadata for the pid from the previous example, then updates 
the access policy that will be applied to the object and uploads the new system metadata to DataONE so that the changes will be applied: 

```{r, eval=F}
cn <- CNode("STAGING")
mn <- getMNode(cn, "urn:node:mnStageUCSB2")
sysmeta <- getSystemMetadata(mn, pid)
sysmeta <- addAccessRule(sysmeta, "public", "read")
status <- updateSystemMetadata(mn, pid, sysmeta)
```
Note that updating an object in DataONE requires the proper access. For example, for the identifier
shown above (urn:uuid:17d61d5a-061a-4778-9cdf-4e14751aaddc), only the DataONE user identity that 
is the *rightsholder*, or another user identity that has been granted write access by the rightsholder
will be able to update the object on the DataONE member node. Running the previous example without
the proper authentication will produce an error.

### Replace an object with a newer version (MNode: updateObject)

The *updateObject* updates an existing object by creating a new object identified by a new PID on the Member Node. The new object replaces and *obsoletes* the old object. An obsoleted object in DataONE does not appear
in search results, however it is still available for download if the identifier is known.

```{r, eval=F}
# Update object from previous example with a new version
updateid <- sprintf("urn:uuid:%s", UUIDgenerate())
testdf <- data.frame(x=1:20,y=11:30)
csvfile <- paste(tempfile(), ".csv", sep="")
write.csv(testdf, csvfile, row.names=FALSE)
size <- file.info(csvfile)$size
sha1 <- digest(csvfile, algo="sha1", serialize=FALSE, file=TRUE)
# Start with the old object's sysmeta, then modify it to match
# the new object. We could have also created a sysmeta from scratch.
sysmeta <- getSystemMetadata(mn, pid)
sysmeta@identifier <- updateid
sysmeta@size <- size
sysmeta@checksum <- sha1
sysmeta@obsoletes <- pid
# Now update the object on the member node.
response <- updateObject(mn, pid, csvfile, updateid, sysmeta)
# Get the new, updated sysmeta and check it to ensure that the update
# worked, i.e. "obsoletes" is the old pid that was replaced by the update.
updsysmeta <- getSystemMetadata(mn, updateid)
updsysmeta@obsoletes
```

The Member Node will mark the object as being *obsolete* by setting a property in the system metadata on the object 
being replaced. An object marked as *obsolete* will not appear in search results, however, such an object is
still available for download if the PID is known.

### Remove an object from DataONE search

An object can be removed from searches done with the DataONE search mechanism by calling the *archive*
method with the PID of the object. This operation does not delete the object bytes, but instead updates
the system metadata for the object to set the *archived* flag to true. The object can still be 
referenced with its PID and downloaded, but it will not appear in any search results.

Objects that are archived can not be updated using the *updateObject* method.  Once an object is archived it cannot be un-archived. 

The following statement archives the object that was just created in the previous example with 
the *updateObject* method.

```{r, eval=FALSE}
response <- archive(mn, updateid)
```

The following commands can be used to verify that the object was archived.
```{r, eval=FALSE}
sysmeta <- getSystemMetadata(mn, updateid)
sysmeta@archived
```